---
title: "Importing and exporting RDES files"
author: "RDML7 authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing and exporting RDES files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(RDML7)
```

## RDES in brief

RDES is the Real-time PCR Data Essential Spreadsheet Format defined by the RDML
Consortium. It is a simpler exchange format for raw qPCR data.

An RDES file:

- represents one run;
- is UTF-8 plain text;
- uses Linux LF newlines;
- is tab-separated even when the extension is `.csv` or `.txt`;
- uses a dot as decimal separator;
- starts with `Well`, `Sample`, `Sample Type`, `Target`, `Target Type`, `Dye`,
  and either `Cq` or `Tm`;
- stores raw fluorescence from column 8 onward;
- stores amplification and melting curves in separate files.

The preferred extension is `.tsv`; `.csv` and `.txt` are also accepted by the
RDES specification.

## Import amplification data

```{r}
rdesamp <- system.file(
  "extdata",
  "RDES_v1_0_example_amplification.tsv",
  package = "RDML7")
amp <- readRDML(rdesamp)
```

The format can also be selected explicitly.

```{r}
amp <- readRDML(
  rdesamp,
  format = "rdes",
  expId = "RDES",
  runId = "run1"
)
```

## Import melting data

```{r}
rdesmelt <- system.file(
  "extdata",
  "RDES_v1_0_example_melting.tsv",
  package = "RDML7")
melt <- readRDML(rdesmelt)

request <- asTable(melt)
request <- request[request$mdp, ]

meltData <- getFData(
  melt,
  request = request,
  dpType = "mdp",
  longTable = TRUE
)
```

## Import an amplification/melting pair

RDES stores the two curve types in independent files. Use `companionFile` to
merge them into one RDML run.

```{r}
both <- readRDML(
  rdesamp,
  format = "rdes",
  companionFile = rdesmelt,
  expId = "exp1",
  runId = "run1"
)

asDendrogram(both)
```

## Column mapping

| RDES | RDML |
|---|---|
| Well | reaction id / position |
| Sample | sample id |
| Sample Type | target-specific sample type |
| Target | target id |
| Target Type | `toi` / `ref` |
| Dye | target dye id |
| Cq | `dataType$cq` |
| Tm | `dataType$meltTemp` |
| cycle columns | amplification `fpoints` |
| temperature columns | melting `fpoints` |

RDES uses `unkn` when the sample type is unavailable and `toi` when the target
type is unavailable.

## Multiple Tm values

RDES allows several Tm values separated by semicolons without spaces:

```text
87.8;82.2
```

The RDML package preserves the full vector in the `meltTemp`. RDES export can
write all values back. Standard RDML XML has a single `meltTemp`, but RDML7
package ignores that!

```{r}
rdesmeltsemi <- system.file(
  "extdata",
  "RDES_v1_0_example_melting_semicolon.tsv",
  package = "RDML7")
meltsemi <- readRDML(
  rdesmeltsemi,
  format = "rdes"
)
meltsemi$experiment$RDES$run$RDES_v1_0_example_melting_semicolon$react$A1$data$`Exon 1`$meltTemp
```

## Strict parsing

*RDES_v1_0_example_melting_semicolon.tsv* contains error: space after *;*.
```{r}
readRDML(
  rdesmeltsemi,
  format = "rdes",
  strict = TRUE
)
```

The normal parser accepts and normalizes some practical text-file deviations;
strict mode converts them into errors.

## Export amplification or melting data

```{r}
writeRDML(
  both,
  "run_amplification.tsv",
  format = "rdes",
  expId = "exp1",
  runId = "run1",
  rdesType = "adp"
)

writeRDML(
  both,
  "run_melting.tsv",
  format = "rdes",
  expId = "exp1",
  runId = "run1",
  rdesType = "mdp"
)
```

## Export both RDES files

```{r}
paths <- writeRDML(
  both,
  "runboth.tsv",
  format = "rdes",
  expId = "exp1",
  runId = "run1",
  rdesType = "both"
)

paths["amplification"]
paths["melting"]
```

The writer creates independent `_amplification` and `_melting` files.

## Raw fluorescence

RDES is intended for raw fluorescence as measured by the instrument. According
to the RDES specification, machine-background correction is acceptable, but
baseline fluorescence correction should not be applied before RDES export.

## Specification

The current RDES description is published by the RDML Consortium at
<https://rdml.org/rdes.html>.
