---
title: "Using the RDML7 package"
author: "Konstantin A. Blagodatskikh, Michał Burdukiewicz, Stefan Rödiger, Andrej-Nikolai Spiess"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the RDML7 package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(RDML7)
```

## Overview

The current `RDML` package uses S7 value objects and a functional camelCase API.
It reads qPCR data, exposes the RDML hierarchy for programmatic editing, provides
tabular fluorescence interfaces, validates semantic consistency, and writes
RDML/RDES and other registered formats.

```{r}
x <- system.file("extdata/stepone_std.rdml", package="RDML7") |> 
  readRDML()

rdmlSummary(x)
validateRDML(x)

metadata <- asTable(x)
amp <- getFData(x, dpType = "adp", longTable = TRUE)

writeRDML(x, "experiment-copy.rdml")
```

Functions that modify an object return a new S7 value:

```{r}
x <- setFData(x, fdata, description)
```

## Reading files

`readRDML()` dispatches through the runtime format registry.

```r
rdmlFormats()
x <- readRDML("experiment.rdml")
```

Use `rdmlDetectFormat()` to inspect selection without performing the import.

```r
rdmlDetectFormat("run.r96")
rdmlDetectFormat("run.tsv")
```

For RDES see `vignette("RDES", package = "RDML")`.

## Navigating the object hierarchy

Schema collections expose virtual keys through `rdmlKeyedList`.

```r
names(x$experiment)

exp <- x$experiment[["exp1"]]
run <- exp$run[["run1"]]
react <- run$react[["A1"]]
data <- react$data[["ACTB"]]
```

The same notation can be used compactly:

```r
x$experiment$exp1$run$run1$react$A1$data$ACTB
```

## Summary and validation

```r
rdmlSummary(x)
issues <- validateRDML(x)
rdmlIsValid(x)
```

Validation can focus on structure, references, curves, or the full object:

```r
validateRDML(x, level = "structure")
validateRDML(x, level = "references")
validateRDML(x, level = "curves")
validateRDML(x, level = "full")
```

## Metadata tables

`asTable()` returns one row per target-specific `dataType` entry.

```r
tab <- asTable(x)
head(tab)
```

Canonical standard columns include:

- `fdataName`;
- `expId`, `runId`, `reactId`;
- `position`;
- `sample`, `sampleType`;
- `target`, `targetDyeId`;
- `adp`, `mdp`.

`namePattern`, `addColumns`, and `...` can evaluate expressions in the current
experiment/run/reaction/data context.

## Fluorescence data

Wide amplification data:

```r
ampWide <- getFData(x, dpType = "adp")
```

Long data joined to metadata:

```r
ampLong <- getFData(
  x,
  dpType = "adp",
  longTable = TRUE
)
```

Melting data:

```r
meltLong <- getFData(
  x,
  dpType = "mdp",
  longTable = TRUE
)
```

In the current implementation, long output is a left join against the supplied
`request`. If a requested entry has no curve of the selected type, its
coordinate and fluorescence fields are `NA`. Filter `asTable()` first when only
actual curves are wanted:

```r
request <- asTable(x)
request <- request[request$mdp, ]

meltLong <- getFData(
  x,
  request = request,
  dpType = "mdp",
  longTable = TRUE
)
```

## Adding fluorescence

```r
fdata <- data.frame(
  cyc = 1:4,
  curve1 = c(10, 12, 20, 50)
)

description <- data.frame(
  fdataName = "curve1",
  expId = "exp1",
  runId = "run1",
  reactId = "A1",
  sample = "sample1",
  sampleType = "unkn",
  target = "ACTB",
  targetDyeId = "FAM"
)

x <- setFData(
  x,
  fdata,
  description,
  fdataType = "adp"
)
```

For constructing a new object from tables use `rdmlFromFData()`. See
`vignette("CreateRDML", package = "RDML")`.

## Merging

```r
merged <- mergeRdmls(
  list(x1, x2),
  dataConflict = "incoming"
)
```

Experiments, runs, reactions, and target-specific data are merged by schema key.

## Writing

```r
writeRDML(x, "result.rdml")
writeRDML(x, "result.xml")
```

Known lossy conversions are controlled with:

```r
writeRDML(x, "result.rdml", loss = "warn")
writeRDML(x, "result.rdml", loss = "error")
writeRDML(x, "result.rdml", loss = "allow")
```

## Legacy API

Compatibility wrappers remain exported for packages using the old API:

```r
AsTable(x)
GetFData(x, long.table = TRUE)
SetFData(x, fdata, legacyDescription, fdata.type = "adp")
```

New code should prefer the camelCase functions.

## Further reading

- `vignette("CreateRDML", package = "RDML")`
- `vignette("RDES", package = "RDML")`
- `vignette("format-plugins", package = "RDML")`
- `vignette("usingRDML_with_other_R_packages", package = "RDML")`
