---
title: "Using the RDML7 package"
author: "RDML7 authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Using the RDML7 package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The **RDML7** package provides tools for importing, exporting, validating,
inspecting, modifying, and converting quantitative PCR (qPCR) data represented
by the Real-time PCR Data Markup Language (RDML).

RDML stores both fluorescence measurements and experimental metadata in a
structured format. The format is designed to make qPCR experiments easier to
exchange, reproduce, and reanalyse.

RDML7 is a major redesign of the original `RDML` R package described in:

> Rödiger S, Burdukiewicz M, Spiess A-N, Blagodatskikh K.
> *Enabling reproducible real-time quantitative PCR research: the RDML package.*
> Bioinformatics 33(24), 4012–4014.
> <https://doi.org/10.1093/bioinformatics/btx528>

The original package used an R6 object model. RDML7 uses **S7 classes** and a
primarily **functional API**. The main workflow is now:

```text
file
  |
  v
readRDML()
  |
  v
rdmlType
  |
  +---- asTable() -------- metadata / curve index
  |
  +---- getFData() ------- fluorescence data
  |
  +---- setFData() ------- add or replace fluorescence data
  |
  +---- rdmlSummary() ---- summary
  |
  +---- validateRDML() --- validation
  |
  +---- asDendrogram() --- hierarchy
  |
  +---- mergeRdmls() ----- merge RDML objects
  |
  +---- writeRDML() ------ RDML / XML / RDES / other registered formats
```

The old names such as `AsTable()`, `GetFData()`, `SetFData()`,
`AsDendrogram()`, `asXML()`, and `MergeRDMLs()` are retained as compatibility
wrappers where possible. New code should use the camelCase API.

# Main changes from the original RDML package

The most important changes are:

1. **S7 object model** instead of R6.
2. **Value semantics**: functions that modify an object return a modified
   object, which should normally be assigned back.
3. **CamelCase functional API**, for example `asTable()` and `getFData()`.
4. **Centralised file I/O** with `readRDML()` and `writeRDML()`.
5. **Extensible format registry** with automatic format detection.
6. **Direct support for importer modules** through a canonical intermediate
   representation.
7. **Validation and summary functions** for RDML objects.
8. **RDES import/export** in addition to RDML and vendor/table formats.
9. A revised `asTable()` interface with documented fields and reusable
   `{field}` name templates.

A migration summary is shown below.

| Original API | RDML7 API |
|---|---|
| `RDML$new(filename = file)` | `readRDML(file)` |
| `RDML$new()` | `rdmlType()` |
| `x$AsTable()` / `AsTable(x)` | `asTable(x)` |
| `x$GetFData()` / `GetFData(x)` | `getFData(x)` |
| `x$SetFData()` / `SetFData(x)` | `x <- setFData(x, ...)` |
| `x$AsDendrogram()` | `asDendrogram(x)` |
| `x$asXML()` | `asXML(x)` or `writeRDML(x, ...)` |
| `MergeRDMLs(...)` | `mergeRDMLs(...)` |
| file-specific constructor logic | `readRDML()` + format registry |

# Loading RDML7

```{r load}
library(RDML7)
```

The package contains example RDML files that can be used throughout this
vignette.

```{r example-file}
filename <- system.file(
  "extdata",
  "lc96_bACTXY.rdml",
  package = "RDML7"
)

filename
```

# Reading files

## `readRDML()`

Use `readRDML()` as the main entry point for file import.

```{r read-rdml}
lc96 <- readRDML(filename)
lc96
```

The default format selection is automatic. If a format is ambiguous, it can be
specified explicitly.

```{r explicit-format, eval=FALSE}
lc96 <- readRDML(
  filename,
  format = "rdml"
)
```

This replaces the old pattern:

```r
# Original RDML package
# lc96 <- RDML$new(filename = filename)
```

## Discovering available file formats

The set of available formats is no longer hard-coded into the user-facing
import function. It is stored in a registry.

```{r formats}
formats <- rdmlFormats()
formats
```

`rdmlFormats()` reports, among other things, the registered format name,
extensions, aliases, whether reading/writing/sniffing is available, the
priority, API version, capabilities, and whether the handler is built in.

To see which reader or writer would be selected for a particular path:

```{r detect-format}
rdmlDetectFormat(
  filename,
  operation = "read"
)
```

RDML7 first uses the extension and, when necessary, registered format sniffers.
If several handlers remain equally suitable, use `format =` explicitly.

The repository includes readers for RDML/XML and several vendor or table
formats. The exact supported set should be queried with `rdmlFormats()` rather
than assumed from a fixed list.

# RDML7 object structure

RDML7 mirrors the hierarchy of the RDML schema. The principal top-level fields
include:

- `dateMade`
- `dateUpdated`
- `id`
- `experimenter`
- `documentation`
- `dye`
- `sample`
- `target`
- `thermalCyclingConditions`
- `experiment`

The important hierarchy containing fluorescence data is approximately:

```text
rdmlType
  `-- experiment
       `-- run
            `-- react
                 `-- data
                      |-- adp   amplification data points
                      `-- mdp   melting data points
```

## S7 property access

RDML7 schema objects are S7 objects. Use `@` to access their properties.

```{r s7-access}
lc96@version

names(lc96@experiment)

experiment <- lc96@experiment[[1]]
experiment@id@id
```

Collections such as experiments, runs, samples, and targets are keyed by their
RDML identifiers. This makes it possible to inspect identifiers first and then
select an element.

```{r keyed-lists}
names(lc96@experiment)

experiment <- lc96@experiment[[1]]

names(experiment@run)
```

For routine data analysis, direct traversal of the full hierarchy is usually
unnecessary. `asTable()` and `getFData()` provide a flatter interface.

# Summarising and validating RDML data

## `rdmlSummary()`

Use `rdmlSummary()` for a compact overview of an RDML object.

```{r summary, eval=FALSE}
rdmlSummary(lc96)
```

This is preferable to manually counting nested experiments, runs, reactions,
targets, and curve types.

## `validateRDML()` and `rdmlIsValid()`

RDML7 provides validation functions in addition to S7 property validators.

```{r validate, eval=FALSE}
validation <- validateRDML(lc96)
validation

rdmlIsValid(lc96)
```

Use validation before exporting data or after programmatically modifying a
large RDML object.

# Visualising the hierarchy

`asDendrogram()` creates a dendrogram representation of the experiment
structure.

```{r dendrogram, eval=FALSE}
asDendrogram(lc96)
```

This is the RDML7 replacement for the old `x$AsDendrogram()` method.

# Creating a table representation with `asTable()`

`asTable()` creates one metadata row per RDML `dataType`. This table is the
main bridge between the hierarchical RDML object and ordinary tabular R
workflows.

```{r as-table}
tbl <- asTable(lc96)

head(tbl)
```

The standard scalar fields are:

- `fdataName` — unique name used to identify a fluorescence series;
- `expId` — experiment identifier;
- `runId` — run identifier;
- `reactId` — reaction identifier;
- `position` — plate/tube position;
- `sample` — sample identifier;
- `target` — target identifier;
- `targetDyeId` — dye associated with the target;
- `sampleType` — sample type;
- `adp` — amplification data are present;
- `mdp` — melting data are present.

The camelCase names replace the old `fdata.name`, `exp.id`, `run.id`,
`react.id`, `sample.type`, and `target.dyeId` columns.

## Selecting standard columns

The revised interface uses `columns` to select built-in output columns.

```{r as-table-columns}
tbl_small <- asTable(
  lc96,
  columns = c(
    "expId",
    "runId",
    "position",
    "sample",
    "sampleType",
    "target"
  )
)

head(tbl_small)
```

Without `columns`, all built-in columns are returned.

## `namePattern` templates

`fdataName` must identify a fluorescence series. RDML7 supports readable
`{field}` templates:

```{r as-table-pattern}
tbl <- asTable(
  lc96,
  namePattern =
    "{expId}_{runId}_{position}_{sample}_{sampleType}_{target}"
)

head(tbl$fdataName)
```

Available built-in template fields are:

```text
{expId}
{runId}
{reactId}
{position}
{sample}
{target}
{targetDyeId}
{sampleType}
{adp}
{mdp}
```

Custom columns supplied through `...` are also available to `namePattern`.

Including `expId` and `runId` is recommended when an object may contain more
than one experiment or run. It substantially reduces accidental duplicate
`fdataName` values.

## Adding custom columns

Custom columns no longer need a separate `addColumns` argument. Supply a named
expression directly through `...`.

```{r custom-column}
tbl <- asTable(
  lc96,
  columns = c(
    "expId",
    "runId",
    "position",
    "sample",
    "target"
  ),
  sampleTarget = paste(
    sample,
    target,
    sep = ":"
  )
)

head(tbl)
```

The ready-to-use scalar fields available in custom expressions are:

```text
expId, runId, reactId, position, sample, target,
targetDyeId, sampleType, adp, mdp
```

### Advanced custom columns

For metadata not exposed as a built-in scalar field, `...` also provides the
current structured RDML objects:

```text
experiment
run
react
data
samples
targets
dateMade
dateUpdated
id
experimenter
documentation
dye
thermalCyclingConditions
```

These objects are intended for advanced extraction. Because they are
structured S7/RDML objects, they are not inserted directly into
`namePattern`.

For example, a scalar property can be extracted from the current `data`
object and then used as a template field:

```{r custom-s7-column, eval=FALSE}
tbl <- asTable(
  lc96,
  cq = S7::prop(data, "cq"),
  namePattern =
    "{expId}_{runId}_{position}_{sample}_{target}_{cq}"
)
```

Once `cq` is created as a scalar custom column, `{cq}` becomes available in
`namePattern`.

This design means users do **not** need to know internal helper functions such
as `.rdmlIdChr()`, `.rdmlReactPosition()`, or `.rdmlSampleType()` merely to
construct an `fdataName`.

## Example: deriving concentration from a sample name

User metadata can also be derived from existing scalar fields. For example,
if a concentration is encoded as a separate underscore-delimited component
of the sample name:

```{r concentration-column, eval=FALSE}
concValues <- c(
  "500", "50", "10", "5", "1",
  "0.5", "0.1", "0.01"
)

tbl <- asTable(
  lc96,
  concentration = {
    parts <- strsplit(
      sample,
      "_",
      fixed = TRUE
    )[[1]]

    value <- parts[
      parts %in% concValues
    ]

    if (length(value)) {
      as.numeric(value[[1]])
    } else {
      NA_real_
    }
  }
)
```

This approach avoids falsely extracting a number that is only part of another
identifier.

## Legacy `default` and `addColumns`

`default` and `addColumns` may be retained for compatibility with older code,
but new code should normally use:

```text
columns + named expressions in ...
```

This gives `asTable()` a single, predictable column-selection model.

# Getting fluorescence data

Fluorescence data are stored separately from the flattened metadata table.
Use `getFData()` to extract them.

## Amplification data (`adp`)

```{r get-adp}
request <- asTable(lc96)

amp <- getFData(
  lc96,
  request = request,
  dpType = "adp"
)

head(amp)
```

For amplification data, the wide table uses `cyc` as its primary coordinate.
If cycle temperatures are present, `tmp` may also be available.

## Melting data (`mdp`)

```{r get-mdp}
melt <- getFData(
  lc96,
  request = request,
  dpType = "mdp"
)

head(melt)
```

For melting data, the coordinate is `tmp`.

## Long-table output

For plotting and grouped statistical analysis, use `longTable = TRUE`.

```{r get-long}
ampLong <- getFData(
  lc96,
  request = request,
  dpType = "adp",
  longTable = TRUE
)

head(ampLong)
```

The long representation combines fluorescence points with the metadata from
the request table.

## Filtering curves before extraction

Because `request` is an ordinary table, it can be filtered before calling
`getFData()`.

```{r filtering}
stdRequest <- request[
  request$sampleType == "std",
]

stdAmp <- getFData(
  lc96,
  request = stdRequest,
  dpType = "adp",
  longTable = TRUE
)

head(stdAmp)
```

This replaces the original workflow in which `AsTable()` output was passed to
`GetFData()`.

## Plotting with ggplot2

```{r plot-curves, eval=FALSE}
library(ggplot2)

ggplot(
  stdAmp,
  aes(
    x = cyc,
    y = fluor,
    group = fdataName,
    colour = target
  )
) +
  geom_line()
```

# Adding or replacing fluorescence data

RDML7 uses **value semantics**. `setFData()` returns the modified object, so
assign it back:

```r
x <- setFData(x, ...)
```

Do not rely on the in-place mutation semantics of the old R6 implementation.

`setFData()` accepts:

- a wide fluorescence table;
- a description table;
- `fdataType = "adp"` or `"mdp"`;
- `conflict = "error"`, `"keep"`, or `"replace"`.

The fluorescence table has a first coordinate column (`cyc` or `tmp`) and one
column per fluorescence series. The series names must match
`description$fdataName`.

# Creating an RDML object from tables

A minimal RDML object can be created with `rdmlType()` and populated with
`setFData()`.

```{r create-from-tables}
fdata <- data.frame(
  cyc = 1:6,
  c1 = c(10, 11, 13, 18, 35, 80),
  c2 = c(9, 10, 12, 16, 29, 65)
)

description <- data.frame(
  fdataName = c("c1", "c2"),
  expId = c("exp1", "exp1"),
  runId = c("run1", "run1"),
  reactId = c("1", "2"),
  sample = c("sample1", "sample2"),
  sampleType = c("unkn", "unkn"),
  target = c("gene1", "gene1"),
  targetDyeId = c("FAM", "FAM"),
  stringsAsFactors = FALSE
)

sim <- rdmlType()

sim <- setFData(
  sim,
  fdata = fdata,
  description = description,
  fdataType = "adp"
)

asTable(sim)
```

The required description columns are:

```text
fdataName
expId
runId
reactId
target
```

Sample and target metadata are used when new schema objects are created.
Missing experiments, runs, reactions, samples, targets, and dyes are created
as needed.

For conflicting existing data:

```{r set-conflicts, eval=FALSE}
sim <- setFData(
  sim,
  fdata,
  description,
  fdataType = "adp",
  conflict = "replace"
)
```

The conflict policies are:

- `"error"` — stop on a conflict;
- `"keep"` — keep existing data;
- `"replace"` — replace existing data.

# Importing vendor and table data

File import is now handled by `readRDML()` and the format registry.

For example:

```{r vendor-import, eval=FALSE}
x <- readRDML("instrument-export-file")
```

If automatic detection is ambiguous:

```{r vendor-import-explicit, eval=FALSE}
x <- readRDML(
  "instrument-export-file",
  format = "registered-format-name"
)
```

Always use:

```{r show-formats-again}
rdmlFormats()
```

to inspect the format names actually registered in the installed version of
RDML7.

The source tree includes importers for RDML/XML, RDES, ABI, Roche,
Rotor-Gene, DTprime, FQD, CSV, Excel/table data, and other supported input
families. The registry is the authoritative runtime list.

# RDES import and export

RDML7 includes RDES support. The example data installed with the package
include amplification and melting RDES TSV files.

```{r rdes-files}
rdesAmp <- system.file(
  "extdata",
  "RDES_v1_0_example_amplification.tsv",
  package = "RDML7"
)

rdesMelt <- system.file(
  "extdata",
  "RDES_v1_0_example_melting.tsv",
  package = "RDML7"
)

c(rdesAmp, rdesMelt)
```

Import uses the same generic entry point:

```{r read-rdes, eval=FALSE}
rdes <- readRDML(rdesAmp)
```

Export is likewise routed through `writeRDML()` according to the requested
output format.

# Merging RDML objects

Use `mergeRdmls()` to combine RDML objects.

```{r merge, eval=FALSE}
file1 <- system.file(
  "extdata",
  "lc96_bACTXY.rdml",
  package = "RDML7"
)

file2 <- system.file(
  "extdata",
  "stepone_std.rdml",
  package = "RDML7"
)

base <- readRDML(file1)
incoming <- readRDML(file2)

merged <- mergeRDMLs(
  base,
  incoming,
  dataConflict = "error"
)
```

`dataConflict` controls what happens when equivalent data locations contain
different values:

- `"incoming"` — use the value from the incoming object;
- `"base"` — preserve the value already present in the base object;
- `"error"` — stop rather than resolve the conflict silently.

For merging independent files, `"error"` is useful while developing a
pipeline because unexpected ID collisions become visible immediately.

A directory of files can be processed in a loop:

```{r merge-directory, eval=FALSE}
files <- dir(
  "path/to/files",
  full.names = TRUE
)

merged <- rdmlType()

for (file in files) {
  current <- readRDML(file)

  merged <- mergeRDMLs(
    merged,
    current,
    dataConflict = "error"
  )
}
```

If experiment identifiers are not unique between input files, rename them
before merging.

# Serialising and writing RDML

## `asXML()`

Use `asXML()` when the XML representation itself is needed:

```{r as-xml, eval=FALSE}
xml <- asXML(lc96)

cat(
  substr(
    xml,
    1,
    500
  )
)
```

## `writeRDML()`

Use `writeRDML()` for normal file output:

```{r write-rdml, eval=FALSE}
writeRDML(
  lc96,
  "lc96-copy.rdml",
  overwrite = TRUE
)
```

For RDML archives, the package writes the XML payload as
`rdml_data.xml` inside the archive.

The writer is selected through the same registry used by `readRDML()`.

# Validation before export

A useful export workflow is:

```{r export-workflow, eval=FALSE}
validation <- validateRDML(lc96)

if (rdmlIsValid(lc96)) {
  writeRDML(
    lc96,
    "validated.rdml",
    overwrite = TRUE
  )
}
```

# Extending the file-format registry

One of the major additions in RDML7 is a public plugin API for file formats.

## Registering a custom format

A format can provide:

- file extensions;
- a reader;
- a writer;
- a sniffer;
- aliases;
- priority;
- API version;
- capabilities.

The public registration function is:

```{r register-format-form, eval=FALSE}
rdmlRegisterFormat(
  name,
  extensions = character(),
  reader = NULL,
  writer = NULL,
  sniff = NULL,
  aliases = character(),
  priority = 0L,
  apiVersion = 1L,
  capabilities = character(),
  overwrite = FALSE
)
```

For example, a third-party package could register a reader:

```{r register-format-example, eval=FALSE}
myReader <- function(fileName, ...) {
  # Parse the external format and return either:
  #   1. an rdmlType object, or
  #   2. an rdmlImportData object.
}

mySniffer <- function(fileName) {
  # Return a confidence score between 0 and 1.
  0
}

rdmlRegisterFormat(
  name = "myformat",
  extensions = "myqpcr",
  reader = myReader,
  sniff = mySniffer,
  aliases = "my-device",
  priority = 10L
)
```

The new format then participates in normal dispatch:

```{r custom-format-dispatch, eval=FALSE}
readRDML("experiment.myqpcr")
```

To remove a third-party format:

```{r unregister-format, eval=FALSE}
rdmlUnregisterFormat("myformat")
```

Built-in handlers are protected from accidental removal.

# Canonical importer API

RDML7 separates **parsing a vendor file** from **constructing the RDML
hierarchy**.

Plugin authors can represent parsed data with:

- `rdmlImportSeries`;
- `rdmlImportData`;
- `rdmlBuildImport()`.

An `rdmlImportSeries` contains:

- `fdataType` (`"adp"` or `"mdp"`);
- a wide fluorescence table accepted by `setFData()`;
- a CamelCase description table accepted by `setFData()`.

`rdmlImportData` groups one or more such series and can contain publisher,
serial number, source format, metadata, and declared conversion losses.

The central builder then constructs `rdmlType`:

```{r canonical-importer, eval=FALSE}
series <- rdmlImportSeries(
  fdataType = "adp",
  fdata = fdata,
  description = description
)

parsed <- rdmlImportData(
  series = list(series),
  publisher = "Example vendor",
  serialNumber = "12345",
  format = "myformat"
)

x <- rdmlBuildImport(parsed)
```

This architecture keeps vendor parsers independent from the nested S7 schema
construction.

# Loss-aware conversions

Different external formats do not necessarily represent every RDML feature.
The importer/exporter API can describe known information loss using
`rdmlLossRecord()`.

When developing a format plugin, declare known lossy transformations rather
than silently discarding unsupported information.

This allows higher-level I/O code to decide whether loss should produce a
warning, an error, or be explicitly allowed.

# `rdmlEdit()`

`rdmlEdit()` remains available as an editing interface.

```{r rdml-edit, eval=FALSE}
rdmlEdit(lc96)
```

For reproducible analyses, scripted operations with `asTable()`,
`getFData()`, `setFData()`, and the S7 object model should remain the primary
workflow, while an editor is useful for interactive inspection and manual
changes.

# Working with external analysis packages

RDML7 intentionally keeps fluorescence extraction separate from downstream
qPCR analysis. Once curves have been obtained with `getFData()`, they can be
passed to other R packages.

For example:

```{r downstream, eval=FALSE}
curves <- getFData(
  lc96,
  dpType = "adp",
  longTable = FALSE
)

# The first column contains cycle coordinates.
cycles <- curves[[1]]

# Remaining columns contain fluorescence series and can be passed to
# modelling, preprocessing, Cq estimation, or visualisation code.
```

This separation allows RDML7 to focus on data interchange, metadata,
validation, conversion, and object manipulation while specialised packages
perform curve analysis.

# Recommended workflow

For most analyses, the following pattern is sufficient:

```{r workflow, eval=FALSE}
library(RDML7)

# 1. Import
x <- readRDML("experiment.rdml")

# 2. Inspect and validate
rdmlSummary(x)
validateRDML(x)

# 3. Build the curve index / metadata table
meta <- asTable(
  x,
  namePattern =
    "{expId}_{runId}_{position}_{sample}_{sampleType}_{target}"
)

# 4. Select curves
selected <- meta[
  meta$sampleType == "std" &
    meta$adp,
]

# 5. Extract fluorescence
amp <- getFData(
  x,
  request = selected,
  dpType = "adp",
  longTable = TRUE
)

# 6. Analyse or modify data
# ...

# 7. Write a new RDML file if needed
writeRDML(
  x,
  "experiment-processed.rdml",
  overwrite = TRUE
)
```

# Legacy API compatibility

Legacy wrappers are exported to ease migration of existing analysis scripts.

Examples include:

```text
AsTable        -> asTable
GetFData       -> getFData
SetFData       -> setFData
AsDendrogram   -> asDendrogram
asXML          -> asXML
MergeRDMLs     -> mergeRdmls
```

Some old dotted argument names are also translated by compatibility code to
their CamelCase equivalents, for example:

```text
show.progress   -> showProgress
conditions.sep  -> conditionsSep
fdata.type      -> fdataType
serial.number   -> serialNumber
name.pattern    -> namePattern
long.table      -> longTable
sample.type     -> sampleType
target.dyeId    -> targetDyeId
dp.type         -> dpType
plot.dendrogram -> plotDendrogram
data.conflict   -> dataConflict
```

Compatibility wrappers are intended to keep old scripts usable. New code and
new documentation should use the canonical names.

# Function overview

The following groups summarise the most important public RDML7 functionality.

## Core object and schema

```text
rdmlType()
idType()
idReferenceType()
experimentType()
runType()
reactType()
dataType()
sampleType()
targetType()
dyeType()
thermalCyclingConditionsType()
...
```

## File I/O

```text
readRDML()
writeRDML()
asXML()
rdmlDetectFormat()
rdmlFormats()
```

## Metadata and fluorescence

```text
asTable()
getFData()
setFData()
rdmlFromFData()
```

## Inspection and validation

```text
rdmlSummary()
validateRDML()
rdmlIsValid()
asDendrogram()
```

## Manipulation

```text
mergeRdmls()
rdmlEdit()
```

## Extensible import/export

```text
rdmlRegisterFormat()
rdmlUnregisterFormat()
rdmlLoadModule()
rdmlImportSeries()
rdmlImportData()
rdmlBuildImport()
rdmlLossRecord()
```

# Summary

RDML7 keeps the main ideas of the original RDML package—structured qPCR
metadata, fluorescence extraction, table-based filtering, creation of RDML
objects from user data, and interoperability with other R analysis
packages—but substantially changes the implementation and public API.

The central concepts for new code are:

1. read files with `readRDML()`;
2. treat `rdmlType` and nested schema elements as S7 objects;
3. use `asTable()` as the metadata/curve index;
4. use `{field}` templates for readable and documented `fdataName` generation;
5. add custom `asTable()` columns directly through `...`;
6. use `getFData()` and `setFData()` for fluorescence;
7. remember that modification uses value semantics;
8. validate with `validateRDML()` / `rdmlIsValid()`;
9. write through `writeRDML()`;
10. discover and extend formats through the format registry.

For the current runtime list of supported file handlers, always prefer:

```r
rdmlFormats()
```

over a hard-coded list in analysis code.
